---
title: "Permutation tests and the Bootstrap"
author: "bioX R-Summer bootcamp"
date: "7/30/2020"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    number_sections: yes
---


# Goals

Let us finish the past sequence of labs on statistical methods (before we delve into the more ML/unsupervised learning side of data analysis) by discussing a very useful class of non-parametric tests: permutation tests and the bootstrap.


After completing this lab, you should be able to...

- Simulate statistics from known distributions to estimate sampling distributions.
- Bootstrap any statistic.
- Create confidience intervals using bootstrap resampling.
- Identify situations in which the bootstrap and permutation tests are appropriate.

Watch the following videos:

- on [permutation tests](https://www.youtube.com/watch?v=GmvpsJHGCxQ)
- on the [bootstrap](https://www.youtube.com/watch?v=9STZ7MxkNVg)

#  The Bootstrap

"The bootstrap is a widely applicable and extremely powerful statistical tool that can be used to quantify the uncertainty associated with a given estimator or statistical learning method." — Page 187, An Introduction to Statistical Learning, 2013.

The bootstrap method is a resampling technique used to estimate statistics on a population by sampling a dataset with replacement. It was proposed by Brad Efron (at Stanford) in 1985, and is an extremely versatile technique (hence, a need-to-know in data analysis), whether it comes to:

- doing hypothesis testing (it's a non-parametric test, so typically when you don't know the distribution of the statistics you're computing, or you don't have enough data)
- computing confidence intervals for your parameters
- finding metrics for assessing the quality of clusters or mixtures (see upcoming discussions.)

It can be used to estimate summary statistics such as the mean or standard deviation. It is used in applied machine learning to estimate the skill of machine learning models when making predictions on data not included in the training data.

Read the section on the [Bootstrap](http://faculty.marshall.usc.edu/gareth-james/ISL/ISLR%20Seventh%20Printing.pdf) p187-190. Once you're done



## Bootstrap for hypothesis testing

Bootstrap tests can be extremely useful for computing intricate null distributions.
The recipe for these is the following:

## Bootstrap for Confidence Interval Estimation



#  Permutation tests

## Exercise 1 - How Large is Large ?
For this exercise we will use:
- Random samples of size n=5, n=25, and n=100.
- Samples from an gamma distribution with
    - $\alpha=0.2$, that is, `shape = 0.2`
    - $\beta=1.4$, that is, `scale = 1.4`
    
Consider using the sample mean, $\bar{x}$, to estimate the mean, $\mu=\mathbb{E}[X]= \alpha \beta=0.28#.

If n is "large" then the central limit theorem suggests that

$$\bar{X} \approx N(\alpha \beta,\frac{\alpha \beta^2}{n})$$

which with some additional work we could then use to create confidence intervals. (We’d also need to estimate the variance.)
However, when is this approximation good?

Perform three simulation studies:

- Study 1: Samples of size n=5
- Study 2: Samples of size n=25
- Study 3: Samples of size n=100

For each, simulate a sample of the specified size from a given gamma distribution 10000 times. For each simulation calculate and store the sample mean.

For each study create a histogram of the simulated sample means. (These will serve as an estimate of the sampling distribution of X¯.) For each, overlay the distribution if the CLT approximation was appropriate:

$$\bar{X} \approx N(\alpha \beta,\frac{\alpha \beta^2}{n})$$

The chunks below outline this procedure.

Hint: Done correctly, you should find that the approximation is bad for n=5, reasonable for n=100 and you may be uncertain about n=25.
```{r}
set.seed(42)
n = 5
sample_means_n_5 = rep(0, 10000)
# perform simulations for n = 5 here
set.seed(42)
n = 25
sample_means_n_25 = rep(0, 10000)
# perform simulations for n = 25 here
set.seed(42)
n = 100
sample_means_n_100 = rep(0, 10000)
# perform simulations for n = 100 here
par(mfrow = c(1, 3))

# create histogram for n = 5 here
# add curve for normal density assuming CLT applies

# create histogram for n = 25 here
# add curve for normal density assuming CLT applies

# create histogram for n = 100 here
# add curve for normal density assuming CLT applies
```

# Permutation test

An increasingly common statistical tool for constructing sampling distributions is the permutation test (or sometimes called a randomization test). Like bootstrapping, a permutation test builds - rather than assumes - sampling distribution (called the “permutation distribution”) by resampling the observed data. Specifically, we can "shuffle" or permute the observed data (e.g., by assigning different outcome values to each observation from among the set of actually observed outcomes). Unlike bootstrapping, we do this without replacement.

Permutation tests are particularly relevant in experimental studies, where we are often interested in the sharp null hypothesis of no difference between treatment groups. In these situations, the permutation test perfectly represents our process of inference because our null hypothesis is that the two treatment groups do not differ on the outcome (i.e., that the outcome is observed independently of treatment assignment). When we permute the outcome values during the test, we therefore see all of the possible alternative treatment assignments we could have had and where the mean-difference in our observed data falls relative to all of the differences we could have seen if the outcome was independent of treatment assignment. While a permutation test requires that we see all possible permutations of the data (which can become quite large), we can easily conduct “approximate permutation tests” by simply conducting a vary large number of resamples. That process should, in expectation, approximate the permutation distribution.

For example, if we have only n=20 units in our study, the number of permutations is:
```{r}
factorial(20)
```

That number exceeds what we can reasonably compute. But we can randomly sample from that permutation distribution to obtain the approximate permutation distribution, simply by running a large number of resamples. Let's look at this as an example using some made up data:
```{r}
set.seed(1)
n <- 100
tr <- rbinom(100, 1, 0.5)
y <- 1 + tr + rnorm(n, 0, 3)
```
The difference in means is, as we would expect (given we made it up), about 1:
```{r}
diff(by(y, tr, mean))
```
To obtain a single permutation of the data, we simply resample without replacement and calculate the difference again:
```{r}
s <- sample(tr, length(tr), FALSE)
diff(by(y, s, mean))
```
Here we use the permuted treatment vector s instead of tr to calculate the difference and find a very small difference. If we repeat this process a large number of times, we can build our approximate permutation distribution (i.e., the sampling distribution for the mean-difference). We'll use replicate do repeat our permutation process. The result will be a vector of the differences from each permutation (i.e., our distribution):
```{r}
dist <- replicate(2000, diff(by(y, sample(tr, length(tr), FALSE), mean)))
```
We can look at our distribution using hist and draw a vertical line for our observed difference:

```{r}
hist(dist, xlim = c(-3, 3), col = "black", breaks = 100)
abline(v = diff(by(y, tr, mean)), col = "blue", lwd = 2)
```

At face value, it seems that our null hypothesis can probably be rejected. Our observed mean-difference appears to be quite extreme in terms of the distribution of possible mean-differences observable were the outcome independent of treatment assignment. But we can use the distribution to obtain a p-value for our mean-difference by counting how many permuted mean-differences are larger than the one we observed in our actual data. We can then divide this by the number of items in our permutation distribution (i.e., 2000 from our call to replicate, above):
```{r}
sum(dist > diff(by(y, tr, mean)))/2000  # one-tailed test
sum(abs(dist) > abs(diff(by(y, tr, mean))))/2000  # two-tailed test
```
Using either the one-tailed test or the two-tailed test, our difference is unlikely to be due to chance variation observable in a world where the outcome is independent of treatment assignment.

### library(coin)
We don't always need to build our own permutation distributions (though it is good to know how to do it). R provides a package to conduct permutation tests called coin. We can compare our p-value (and associated inference) from above with the result from coin:
```{r}
library(coin)
independence_test(y ~ tr, alternative = "greater")  # one-tailed
```

```{r}
independence_test(y ~ tr)  # two-tailed
```

Clearly, our approximate permutation distribution provided the same inference and a nearly identical p-value. coin provides other permutation tests for different kinds of comparisons, as well. Almost anything that you can address in a parametric framework can also be done in a permutation framework (if substantively appropriate). and anything that coin doesn't provide, you can build by hand with the basic permutation logic of resampling.

