---
title: "Week 1 Session 2"
subtitle: "Basics of Coding in R"
date: July 1st, 2020
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# Watch the video for Session 2

Watch the video lecture associated to Session 2 (find the link on the website) -- it will walk you through some of the basic syntax in R.
Once you're done, go through the following exercises to get more familiar with the R syntax.


# Exercise 0: Swirl courses

The purpose of this first lab is to familiarize yourself with the R language. 
To this end, we will use one of the Swirl courses (which we installed during the discussion session).

Installation:

```{r cars, eval=FALSE}
library(swirl)
install_course_github("swirldev", "R_Programming_E")
swirl()
```

This will initiate a series of promts:

- Type 1 for "1: R Programming: The basics of programming in R" when prompted to install a course.
- Type 1 for "1: R Programming" when prompted to choose a course.

Go through lessons 1 through 11.
Once you're done, prepare the following exercises which we will go over during the discussion.

# Exercise 1: Vectors

1. Generate and print a vector of 10 random numbers between 5 and 500.
2. Generate a random vector Z of 1000 letters (from "a" to "z"). Hint: the 
variable `letters` is already defined in R.
3. Print a summary of Z in the form of a frequency table.
4. Print the list of letters that appear an even number of times in Z.

# Exercise 2: Matrices

1. Create the following 5 by 5 matrix and store it as variable X.
```{r echo = FALSE}
matrix(1:30, nrow = 5, byrow = FALSE)
```

2. Create a matrix Y by adding an independent Gaussian noise (random numbers)
with mean 0 and standard deviation 1 to each entry of X. e.g.

3. Find the inverse of Y.

4. Show numerically that the matrix product of Y and its inverse 
is the identity matrix.


# Exercise 3: Data frames

1. Create the following data frame and name it “exams”.

```{r, echo = FALSE}
set.seed(123)
data.frame(
  student = c("Alice", "Sarah", "Harry", "Ron", "Kate"),
  score = sample(80:100, 5),
  letter = sample(c("A","B"), 5, replace = TRUE),
  late = sample(c(T, F), 5, replace = TRUE)
)
```

2. Compute the mean score for this exam and print it.
3. Find the student with the highest score and print the corresponding 
row of "exams". Hint: use the function `which.max()`.


# Execise 4: Control Flow 

## Part 1.1

Use a for loop to:

a. Print all the letters of the Latin alphabet:`

b. Print the numbers 10 to 100 that are divisible by 7

c. Print the numbers 1 to 100 that are divisible by 5 but not by 3.



## Part 1.2

a. Find all numbers not greater than 10,000 that are divisible by 
5, 7 and 11 and print them.

b. Print for each of the numbers x = 2, . . . 20, all numbers that divide x
(all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.


# Execise 5: Functions

## Part 2.1

a.Create a function what will return the number
of times a given integer is contained a given vector of integers.
The function should have two arguments one for a vector
and the other for a scalar. 

b. Then, generate a random vector of 100 
integers (in a range 1-20) use the function to count the number
of times the number 12 is in that vector. 

## Part 2.2

Write a function that takes in a data.frame as an input, 
prints out the column names, and returns its dimensions.

# Execise 6: Apply family functions

## Part 1
Below we print six first rows of the built-in dataset, `mtcars`, 
from the 1974 Motor Trend US magazine, which comprises information
on the fuel consumption and 10 aspects of automobile design 
and performance for 32 selected car models. 

```{r}
head(mtcars)
```

Use `apply()` function to find the standard deviation and the 0.8-quantile  
of each of the automobile characteristic.

## Part 2

Below is a vector of dates in year 2018.

```{r}
set.seed(1234)
y2018 <- seq(as.Date("2018-01-01", format = "%Y-%m-%d"), 
             as.Date("2018-12-31", format = "%Y-%m-%d"), 
             "days")
length(y2018)
# A random sample of 10 dates from 2018
y2018_sample <- sample(y2018, size = 10)
y2018_sample
```

Use an `apply` family function to return the number of weeks
left from each day in `y2018_sample` to the New Year, 2019/01/01.


