---
title: "Lab 2"
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    number_sections: no
    toc: yes
    toc_float: true
    toc_depth: 3
    self_contained: yes
date: "July 1st, 2020"
subtitle: Basics of Coding in R
---

# Watch the video series for Session 2

Watch the video lecture associated to Session 2 (find the link on the website and [here](https://www.youtube.com/watch?v=iffR3fWv4xw&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP)) -- these are a series of short videos that will walk you through some of the basic syntax in R. In particular, we ask that you watch the videos in that link entitled:

- R [1.1](https://www.youtube.com/watch?v=iffR3fWv4xw&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=1), [1.2](https://www.youtube.com/watch?v=S-o-sdlzhkE&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=2), [1.3](https://www.youtube.com/watch?v=YhQOV27pQfg&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=3), [1.4](https://www.youtube.com/watch?v=GKu5tw_bIpA&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=4), [1.5](https://www.youtube.com/watch?v=u5hroyx0J4o&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=5), [1.6](https://www.youtube.com/watch?v=cR-hEUs1rRw&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=6), and [1.7](https://www.youtube.com/watch?v=cR-hEUs1rRw&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=7)

- R [2.1](https://www.youtube.com/watch?v=qK1ElUMkhq0&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=8), R [2.3](https://www.youtube.com/watch?v=eVEx_pBEkRI&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=10) and R [2.4](https://www.youtube.com/watch?v=djI-yfk-DZM&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=11)

- R [3.2](https://www.youtube.com/watch?v=WuCVs3bW-ZY&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=14)

- R [4.1](https://www.youtube.com/watch?v=Z1wB1rHAYzQ&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP&index=16)

Once you're done, go through the following exercises to get more familiar with the R syntax.


# Exercise 0: Swirl courses

The purpose of this first lab is to familiarize yourself with the R language. 
To this end, we will use one of the Swirl courses (which we installed during the discussion session).

Installation:

```{r cars, eval=FALSE}
library(swirl)
install_course_github("swirldev", "R_Programming_E")
swirl()
```

This will initiate a series of prompts:

- Type 1 for "1: R Programming: The basics of programming in R" when prompted to install a course.
- Type 1 for "1: R Programming" when prompted to choose a course.

Go through lessons 1 through 11.

Once you're done, prepare the following exercises which we will go over during the discussion.

# Exercise 1: Vectors

1. Generate and print a vector of 10 random numbers between 5 and 500.
2. Generate a random vector Z of 1000 letters (from "a" to "z"). Hint: the variable `letters` is already defined in R.
3. Print a summary of Z in the form of a frequency table.
4. Print the list of letters that appear an even number of times in Z.

# Exercise 2: Matrices

1. Create the following 5 by 5 matrix and store it as variable X.
```{r echo = FALSE, eval = TRUE}
matrix(1:25, nrow = 5, byrow = FALSE)
```

2. Create a matrix Y by adding an independent Gaussian noise (random numbers) with mean 0 and standard deviation 1 to each entry of X. Hint: use the `rnorm` function.

3. Find the inverse of Y. Hint: use the `solve` function.

4. Show numerically that the matrix product of Y and its inverse is the identity matrix. Hint: use the `%*%` matrix multiplication operator.


# Exercise 3: Data frames

1. Create the following data frame and name it “exams”.

```{r, echo = TRUE}
set.seed(123)
data.frame(
  student = c("Alice", "Sarah", "Harry", "Ron", "Kate"),
  score = sample(80:100, 5),
  letter = sample(c("A","B"), 5, replace = TRUE),
  late = sample(c(T, F), 5, replace = TRUE)
)
```

2. Compute the mean score for this exam and print it.

3. Find the student with the highest score and print the corresponding row of "exams". Hint: use the function `which.max()`.


# Exercise 4: Control Flow 

## Part 1

Use a "for"" loop to:

a. Print all the letters of the Latin alphabet.

b. Print the numbers 10 to 100 that are divisible by 7

c. Print the numbers 1 to 100 that are divisible by 5 but not by 3.



## Part 2

a. Find all numbers not greater than 10,000 that are divisible by 5, 7 and 11 and print them.

b. Print for each of the numbers x = 2, . . . 20, all numbers that divide x (all factors) excluding 1 and x. Hence, for 18, it should print 2 3 6 9.


# Exercise 5: Functions

## Part 1

a. Create a function what will return the number of times a given integer is contained a given vector of integers. The function should have two arguments one for a vector and the other for a scalar. 

b. Then, generate a random vector of 100 integers (in a range 1-20) use the function to count the number of times the number 12 is in that vector. 

## Part 2

Write a function that takes in a data.frame as an input, prints out the column names, and returns its dimensions.

# Exercise 6: Apply family functions

## Part 1
Below we print six first rows of the built-in dataset, `mtcars`, from the 1974 Motor Trend US magazine, which comprises information on the fuel consumption and 10 aspects of automobile design and performance for 32 selected car models. 

```{r}
head(mtcars)
```

Use `apply()` function to find the standard deviation and the 0.8-quantile of each of the automobile characteristic.

## Part 2

Below is a vector of dates in year 2018. Hint: you might find the `ceiling` function useful.

```{r}
set.seed(1234)
y2018 <- seq(as.Date("2018-01-01", format = "%Y-%m-%d"), 
             as.Date("2018-12-31", format = "%Y-%m-%d"), 
             "days")
length(y2018)
# A random sample of 10 dates from 2018
y2018_sample <- sample(y2018, size = 10)
y2018_sample
```

Use an `apply` family function to return the number of weeks left from each day in `y2018_sample` to the New Year, 2019/01/01.

Note: you can calculate the difference between Date objects.

```{r}
ny2019 <- as.Date("2019-01-01", format = "%Y-%m-%d")
ny2018 <- as.Date("2018-01-01", format = "%Y-%m-%d")
ny2019-ny2018
```


