---
title: 'Lab 5b: Solutions to Mixture Models and Variance Stabilization'
author: "Bios221"
date: "10/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Question 1:** How many modes are there?

**Answer:** There are 2 modes.

**Question 2:** Modify the code below to simulate one million coin flips 
and make a histogram with 500 bins.

```{r}
library(ggplot2)

sd1 = 0.5 ; sd2 = 0.5
mean1 = 1 ; mean2 = 3

B = 1e+6
coinflips = (runif(B)>0.5)
fairmix = rnorm(length(coinflips),
                mean = ifelse(coinflips, mean1, mean2),
                sd   = ifelse(coinflips, sd1,   sd2))
fairdf = data.frame(fairmix)
# Make the plot below using bins = 500 instead of binwidth = 0.2 in geom_histogram()

ggplot(fairdf, aes(x = fairmix)) +
    geom_histogram(fill = "red", color = "grey", bins = 500)
```


**Question 3**:
a) Try to increase the samples size n1 and n2. What happens then?
b) Why do you get different results every time you run this code?
c) How can you make it produce always the same result?

**Answer:**
a) As n1, n2 increase, we can learn the true means quite accurately.
b) The EM algorithm is initialized from random centers every time, read the documentation to `?normalmixEM`.
c) By setting a seed using e.g., `set.seed(1)`.



**Question 4**:Give the upper end of a centered 95\% confidence interval for the median based on the above bootstrap sample. Tip: Use the ``quantile`` function.

**Answer:**
```{r}
library("HistData")

set.seed(1)
B = 10000
diff = ZeaMays$diff
samplesIdx = replicate(B, sample(15, 15, replace = T))
samplingDist = apply(samplesIdx, 2, function(x) {return(median(diff[x]))})

quantile(samplingDist, 0.975)
```